Wind veer and and Speed in Turbulent Ekman Flow

### 2022-01-11 write1/manuscript.tex
further work on manuscript

\date{\scriptsize Draft - Version \today}
\title{Profiles of Wind Direction and Speed in Turbulent Ekman Flow}
\author{Cedrick Ansorge, Hauke Wurps}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
At the small-scale end, the directional shear of the wind in the upper part of the surface layer
may cause a systmatic yaw for tall wind power generation devices where blades reach into the Ekman layer,
i.e. that part of the boundary layer where the wind starts to turn. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer and a first
attempt in theoretical quantification was undertaken by \citet{rossby:PPO1935} using integral relations in the boundary layer.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}. 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

@@

@@

For instance, effects of roughness are commonly incorporated by a linear transformation of vertical scale involving the

@@

effects of stability are commonly accounted for by a linearization around the neutrally stratified profile~%
\citep{monin:ARF1970, monin:1975, hogstrom:BM1988, hogstrom:BM1996,sakagami:BM2020}.
Furthermure, profiles can serve as better initial condition for numerical simulation of the flow,

@@

this conceptually well-defined problem.

@@

In fully developed turbulence certain statistics such as dissipation \citep{dimotakis:ARF2005} or profiles of
mean velocity \citep{barenblatt:JFM1993} become independent of the Reynolds number when scaled appropriately while
other statistics, such as the near-wall maximum in velocity fluctuation depend on $\RE$ \citep{baars:JFM2020a} and external
interfaces may exert an impact on near-wall scaling \citep{dasilva:ARF2014}. 

@@

Reynolds numbers that became possible with the increasing capabilities of supercomputering over

@@

\section{Problem Formulation and Numerical Approach}

>>>>

\date{\scriptsize Draft - Version \today}
\title{Wind veer and and Speed in Turbulent Ekman Flow}
\author{Cedrick Ansorge, Hauke Wurps}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales \todo{Insert: Brown, 2005; Svensson and Holtslag, 2009}.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systmatic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\todo{References: Calaf et al., 2010 Phys Fluids; Mirocha et al., 2018 Wind Energy Science} . 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer
and a \citet{rossby:PPO1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
\todo{Refer to Pena et al, 2010} 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

@@

More recently, \citet{ghannam:jas2021} treated the horizontally averaged momentum budget to show that departures from
shear-alignment in the vicinity of the surface result in an integral of the wind veer ($\alpha_M$ in their notation)
over the height to very high accuracy ($\int_{z_0}^{H} \sin\alpha_M$ in their notation; their Eq.~(16)).
Classic surface-layer similarity is recovered when the angle $\alpha_M$ does not depend on height, i.e., the
wind veer is constant across the surface layer. 
If, however, the wind veer depends on height, the profiles of stress and mean velocities depart from the
scalings implied by classic surface-layer similarity.

@@

More complex accounts generally refer to the homogeneous stationary problem as a base state: 
(i) Roughness is commonly incorporated by a linear transformation of vertical scale involving the

@@

(ii) Stability can be accounted for by a linearization around the neutrally stratified profile~%
\citep{monin:ARF1970, monin:1975, hogstrom:BM1988, hogstrom:BM1996,sakagami:BM2020};
(iii) Non-stationarity in the pressure-gradient forcing can be accounted for by a linear damped-oscillator
approach around the base state\citep{momen:jas2016}; 
(iv) Barotropic and baroclinic effects on the velocity profile require to consider 
the height-dependence of the veer and stress misalignment \citep{momen:jas2018, ghannam:jas2021}.
Furthermore, such a solution can serve as better initial condition for numerical simulation of the flow,

@@

this well-defined problem.

@@

Certain statistics of fully developed turbulence, such as dissipation \citep{dimotakis:ARF2005} or profiles of
mean velocity \citep{barenblatt:JFM1993}, become independent of the Reynolds number when appropriately scaled; 
other statistics, such as the near-wall maximum in velocity fluctuation depend on $\RE$ \citep{baars:JFM2020a} and externality
of the flow may exert an impact on near-wall scaling \citep{dasilva:ARF2014}. 

@@

Reynolds numbers that became possible due to an increase of computing capabilities over

@@

\section{Problem Form
  ulation and Numerical Approach}
